It still seems that black has hope in these extremely unfair variants of chess

### 2014-03-17 chesstego/paper.tex
Submitted version, delete unused chesswho

\title{It still seems that black has hope in these extremely unfair variants of chess}
\author{Dr.~Tom~Murphy~VII~Ph.D. \and Ben Blum \and
Jim McCann, in italics ...... oh! .... nervous laughter .... no, that's not part of the name still. I stopped saying the name. Why are you still typing? This isn't the name any more. End quote.\thanks{
Copyright \copyright\ 2014 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2014 with the undivided
attention of the Association for Computational Heresy; 
{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
\$0.00}}

@@

In Stratego, each player begins the game by arranging his or her
Stratego-pieces in a secret fashion on the board. In the world of
Stratego, civization is ruled by a leader known as Flag. The player's
goal is to assassinate the opponent Flag, using a member of his or her
army. However, which piece represents Flag is unknown!

@@

\paragraph{Chesstego variation I.} In this variation, each player decides
in secret, before the game begins, which of his or her pieces is the
actual King, that is, the Flag. If this piece is checkmated, the game
is lost. There are two subvariations: {\em I(a)}, where a player must
announce {\em Check!} when the Flag is under attack, and all of the
normal rules about moving into (or castling through) check must be
obeyed. In subvariation {\em I(b)}, which I prefer, the piece that is
Flag may slip silently into and out of check, and the game only ends
when that piece is captured.\footnote{In Chess proper, these
  formulations are nearly equivalent---except for rules like castling
  through check and some corner cases---but it is deemed important for
  movie drama that players be able to announce {\em Check!} and {\em
    Checkmate!} at times.}

@@

\paragraph{Chesstego variation II.} In this variation, each player
decides in secret, before the game, which of her {\em opponent's} pieces
is Flag. Same as before, if this piece is captured, the game ends
instantly. But now, players don't even know which of their own pieces
is their glorious ruler, Flag. This can be very exciting or tittilating.

@@

\chessboard[
  setfen=4k3/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR]

Lo! This version is finally satisfactory. In this version we need not
equivocate over who shall win. White picks one of her very safe pieces
(e.g., one of the rooks) as Flag, and chooses Black's b7~pawn as
Black~Flag. White's winning move is c2c4 and then Qb3, with Black~Flag
unable to escape the B file. Choosing the f7~pawn works as well.

@@

Example hopeless match: 1. c4 Kf8~~2. Qb3 b5~~3. c4xb4\checkmate \\
\chessboard[setfen=5k2/p1pppppp/8/1P6/8/1Q6/PP1PPPPP/RNB1KBNR]

@@

\chessboard[
  setfen=4k3/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]

@@

In {\em Chess Who?}, which is the chess version, players have two
different options on each turn. They can either move a piece like
normal, or ask a question of their opponent. The question must be
about the physical characteristics of the pieces, for example, ``Does
your piece have curly hair?'' No pieces have curly hair, so nothing
happens. The opponent might ask a different question, on the next turn,
like ``Does your piece have straight hair?'' No pieces have straight
hair, so nothing happens. On the next turn, suppose the player asks,
``Does your piece have brown hair?'' No pieces have brown hair,
nor any hair at all, and no pieces have any brown parts at all either.
Stop asking. Next, Black asks, ``Does your piece have crennelations?''
Both of White's rooks have crennelations, so they are eliminated from
the board (Figure~\ref{fig:crennelations}).

\begin{figure}[ht]
\begin{center}
\chessboard[
   setfen=rnbqkbnr/pppppppp/8/8/8/8/PPPPPPPP/1NBQKBN1]
\end{center}\vspace{-0.1in}
\caption{
  The board after
  1. Does your piece have curly hair?
     \ldots Does your piece have straight hair?
  2. Does your piece have brown hair?
     \ldots Does your piece have crennelations?
}
\label{fig:crennelations}
\end{figure}

@@

This variation produces draws easily. An example Nash Equilibrium:
White begins by asking, ``Does your piece have three lobes like a
truncated snowman, {\em or} crennelations, {\em or} horse ears, {\em
  or} two weird long shoulder-ribbons that I don't know what they are,
{\em or} five pointy tines on her crown?'' This matches all of Black's
pieces except his king, so he removes them all. But then Black asks:
``Does your piece {\em not} have a lumpy crown?'' which matches all of
White's pieces except his king. So all we have left is the two kings,
which is known to be a draw~\cite{pawnless}.

@@

{\em Chessy Crush Saga} is a hybrid of Chess and the popular and
lucrative phone-game {\em Candy Crush Saga}. In this game, the
regular moves of chess are allowed, but it is also possible to
{\em crush} chess pieces into candies. {\em Chessy Crush v1}
follows the rules of {\em Candy Crush} closely: When a piece moves
such that three pieces of the same color are in a row (either
horizontally or vertically), those pieces are all removed. Usually
this is a tactical misstep in Chess, as three of one's own pieces
are destroyed for nothing. However, as in {\em Candy Crush}, rewards
are given when more than three like pieces are destroyed at once.
The simplest is to award the player with a Candy Rook at at the
place where the moved piece ended up. A Candy Rook moves like a
rook, but if it is crushed, it destroys all of the opponent's
pieces on the file (column) it currently occupies.\!\footnote{In
formal {\em Candy Crush}, this is the vertical striped candy,
which is awarded when the piece was moved vertically. To keep
this variant simple, we always award a ``vertically striped''
Candy Rook, since it seems with only pawns, only vertical moves
are likely. (This later turns out to be a bad assumption.)}

@@

\begin{figure}[ht]
\begin{center}
\chessboard[setfen=4k3/1ppppppp/8/8/8/1p1PP3/P1P2PPP/4K3]
\end{center}\vspace{-0.1in}
\caption{
  Black interfering with White's Omega Weapon main line:
  b3, d3, e3, c3 {\em (making candy)}, Rc2, Re2++. Black
  can progress pawns fast enough to capture White's. She
  can also move her King, making the candied rook less
  devastating. Ultimately, White's one-tempo advantage
  is not obviously winning.
}
\label{fig:chessyv1}
\end{figure}

In {\em v2}, the Candy Rook is extremely powerful, since
not only does it destroy an entire file (the e file being
the likely target, since it contains the immovable King),
but when it is used (by crushing) it is also replaced,
since a crush always yields candy. Therefore, White's
most straightforward strategy is to quickly create a
Candy Rook, then crush it in the e file, which wins. This
should be a line where White's tempo advantage makes it
mostly immune from interference, since even a check is
answered by destroying the opponent's king.

% \chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
% \chessboard[setfen=4k3/ppp1pppp/8/8/8/4p3/PP3PPP/4K3]
% \chessboard[setfen=4k3/1ppppppp/8/8/8/8/pR2PPPP/4K3]
\begin{figure}[ht]
\begin{center}
\chessboard[setfen=4k3/3ppppp/1r6/8/8/2R5/PP3PPP/4K3]
(a) \\
\chessboard[
  setfen=4k3/3ppppp/4r3/8/8/8/2R2PPP/4K3,
  pgfstyle=border,
  color=blue!50,
  markfile=e,
  color=black,
  pgfstyle=straightmove,
  markmoves={b6-e6},
  pgfstyle=cross,
  color=red,
  backfields={e8,e7,e6}]
(b)
\end{center}\vspace{-0.1in}
\caption{ 
%
  Black's surprising draw strategy in {\em Chessy Crush v2}. White
  begins her Omega Weapon: 1. e3 c6 2. d3 a6 3. c3 {\em making candy}
  b6 {\em making candy} we have the board in (a). It seems that Black
  is stuck after 4. Rc2, since 5. Re2 will crush White's candy rook
  and destroy the entire e file, including Black's king. Black
  defending with 4. \ldots Re6 initially seems pointless (White will
  destroy the rook as well so the check is irrelevant---this indeed
  refutes 4. \ldots Rb1) but Re6 actually creates a {\em vertical}
  three-piece crush, sacrificing Black's king to destroy the e file
  first, which also destroys White's king! Such mutual destruction
  is not possible in traditional chess, but it seems appropriate to
  consider it a draw in Chessy Crush Saga v2. It appears that White
  can head off this line with 4. Rc8++, a totally vanilla back-rank mate,
  but this is not actually mate with the same 4. \ldots Rb1 escaping
  to a draw.
%
}
\label{fig:chessyv2}
\end{figure}

% Chessy Crush Saga
% Battleship + Chess = Battlechess
% Chess Tac Toe
% Chesslers of Catan
% Chesstris
% Hungry Hungry Hipposchess

% Monopoly?

\setchessboard{tinyboard,showmover=false}
% \chessboard[
%   setfen=rnbqkbnr/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]

@@

>>>>

\title{It still seems that black has hope in these extremely unfair variants of chess}
\author{Dr.~Tom~Murphy~VII~Ph.D. \and Ben Blum \and
Jim McCann, in italics ...... oh! .... nervous laughter .... no, that's not part of the name still. I stopped saying the name. Why are you still typing? This isn't the name any more. End quote.\thanks{
Copyright \copyright\ 2014 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2014 with the undivided
attention of the Association for Computational Heresy; 
{\em IEEEEEE!} press, Verlag-Verlag volume no.~0x40-2A.
\$0.00}. Badness: 12,844 }

@@

In Stratego, each player begins the game by arranging his or her
Stratego-pieces in a secret fashion on the board. In the world of
Stratego, civilization is ruled by a leader known as Flag. The
player's goal is to assassinate the opponent Flag, using a member of
his or her army. However, which piece represents Flag is unknown!

@@

\paragraph{Chesstego variation I.} In this variation, each player decides
in secret, before the game begins, which of his or her pieces is the
actual King, that is, Flag. If this piece is checkmated, the game
is lost. There are two subvariations: {\em I(a)}, where a player must
announce {\em Check!} when Flag is under attack, and all of the
normal rules about moving into (or castling through) check must be
obeyed. In subvariation {\em I(b)}, which I prefer, the piece that is
Flag may slip silently into and out of check, and the game only ends
when that piece is captured.\!\footnote{In Chess proper, these
  formulations are nearly equivalent---except for rules like castling
  through check and some corner cases---but it is deemed important for
  movie drama that players be able to announce {\em Check!} and {\em
    Checkmate!} at times.}

@@

\paragraph{Chesstego variation II.} In this variation, each player
decides in secret, before the game, which of her {\em opponent's} pieces
is Flag. Same as before, if this piece is captured, the game ends
instantly. But now, players don't even know which of their own pieces
is their glorious ruler, Flag. This can be very exciting or titillating.

@@

\begin{center}
\chessboard[
  smallboard,
  setfen=4k3/pppppppp/8/8/8/8/PPPPPPPP/RNBQKBNR]
\end{center}

Lo! This version is finally satisfactory. In this version we need not
equivocate over who shall win. White picks one of her very safe pieces
(e.g., one of the rooks) as Flag, and chooses Black's b7~pawn as
Black~Flag. White's winning move is c2c4 and then Qb3, with Black~Flag
unable to escape the B file~(Figure~\ref{fig:hopeless}). Choosing the f7~pawn works as well.

@@

\begin{figure}[ht]
\begin{center}
\chessboard[setfen=5k2/p1pppppp/8/1P6/8/1Q6/PP1PPPPP/RNB1KBNR,
  pgfstyle=straightmove,
  markmoves={c4-b5}
]
\end{center}\vspace{-0.1in}
\caption{
Example hopeless match: 1. c4 Kf8~~2. Qb3 b5~~3. xb5 1--0
}
\label{fig:hopeless}
\end{figure}

@@

\begin{center}
\chessboard[
  smallboard,
  setfen=4k3/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]
\end{center}

@@

In {\em Chess Who?}, which is the chess version, players have two
different options on each turn. They can either move a piece like
normal, or ask a question of their opponent. The question must be
about the physical characteristics of the pieces, for example, ``Does
your piece have curly hair?'' No pieces have curly hair, so nothing
happens. The opponent might ask a different question, on the next turn,
like ``Does your piece have straight hair?'' No pieces have straight
hair, so nothing happens. On the next turn, suppose the player asks,
``Does your piece have brown hair?'' No pieces have brown hair,
nor any hair at all, and no pieces have any brown parts at all either.
Stop asking. Next, Black asks, ``Does your piece have crenellations?''
Both of White's rooks have crenellations, so they are eliminated from
the board (Figure~\ref{fig:crenellations}).

\begin{figure}[ht]
\begin{center}
\chessboard[
   setfen=rnbqkbnr/pppppppp/8/8/8/8/PPPPPPPP/1NBQKBN1]
\end{center}\vspace{-0.1in}
\caption{
  The board after
  1. Does your piece have curly hair?
     \ldots Does your piece have straight hair?
  2. Does your piece have brown hair?
     \ldots Does your piece have crenellations?
}
\label{fig:crenellations}
\end{figure}

@@

This variation produces draws easily. An example Nash Equilibrium:
White begins by asking, ``Does your piece have three lobes like a
truncated snowman, {\em or} crenellations, {\em or} horse ears, {\em
  or} two weird long shoulder-ribbons that I don't know what they are,
{\em or} five pointy tines on her crown?'' This matches all of Black's
pieces except his king, so he removes them all. But then Black asks:
``Does your piece {\em not} have a lumpy crown?'' which matches all of
White's pieces except his king. So all we have left is the two kings,
which is known to be a draw~\cite{pawnless}.

@@

{\em Chessy Crush Saga} is a hybrid of Chess and the popular and
lucrative phone-game {\em Candy Crush Saga}. In this game, the
regular moves of chess are allowed, but it is also possible to
{\em crush} chess pieces into candies. {\em Chessy Crush v1}
follows the rules of {\em Candy Crush} closely: When a piece moves
such that three pieces of the same color are in a row (either
horizontally or vertically), those pieces are all removed. Usually
this is a tactical misstep in Chess, as three of one's own pieces
are destroyed for nothing. However, as in {\em Candy Crush}, rewards
are given when more than three like pieces are destroyed at once.
The simplest is to award the player with a Candy Rook at the
place where the moved piece ended up. A Candy Rook moves like a
rook, but if it is crushed, it destroys all of the opponent's
pieces on the file (column) it currently occupies.\!\footnote{In
formal {\em Candy Crush}, this is the vertical striped candy,
which is awarded when the piece was moved vertically. To keep
this variant simple, we always award a ``vertically striped''
Candy Rook, since it seems with only pawns, only vertical moves
are likely. (This later turns out to be a bad assumption.)}

@@

With just pawns, it seems that the best way to gain advantage for
White is to crush four pawns to create a Candy Rook before Black does,
then menace Black with it. Unfortunately, moving four pawns into
formation before black can interfere is not possible; Black can react
to White's first move and advance one of his own pawns towards the
construction (Figure~\ref{fig:chessyv1}). Moreover, if he does, and
trades for one of White's pawns (or better, the resulting candy), then
Black is probably in a superior position, White having crapified his
pawn structure in pursuit of candy. Worse still, crushing a successful
Candy Rook is not that menacing, since Black can simply move his king
out of the way.

\begin{figure}[ht]
\begin{center}
\chessboard[setfen=4k3/1ppppppp/8/8/8/1p1PP3/P1P2PPP/4K3,
  pgfstyle=straightmove,
  markmoves={a4-b3},
]
\end{center}\vspace{-0.1in}
\caption{
  Black interfering with White's Omega Weapon main line:
  b3, d3, e3, c3 {\em (making candy)}, Rc2, Re2++. Black
  can progress pawns fast enough to capture White's. She
  can also move her King, making the candied rook less
  devastating. Ultimately, White's one-tempo advantage
  is not obviously winning.
}
\label{fig:chessyv1}
\end{figure}

To correct these problems with balance ({\em i.e.}, that the game may
be balanced), we simplify the candying rules: Any crush produces a
Candy Rook, including ones of length 3. This allows White to create a
Candy Rook before Blacks pawns can reach her. Furthermore, we
stipulate that the kings cannot move (they don't have legs anyway,
this is plain to see, so let's be realistic). In {\em v2}, the Candy
Rook is extremely powerful, since not only does it destroy an entire
file (the e file being the likely target, since it contains the
immovable King), but when it is used (by crushing) it is also
replaced, since a crush always yields candy. Therefore, White's most
straightforward strategy is to quickly create a Candy Rook, then crush
it in the e file, which wins. This should be a line where White's
tempo advantage makes it mostly immune from interference, since even a
check is answered by destroying the opponent's king.

The idea behind White's Omega Weapon is to create a triplet of pawns
on rank 3, either c3--e3 or e3--g3. White begins by moving her king's
pawn to e3. Black can interfere with one side of the triplet
(Figure~\ref{fig:twopaths}), but only one. After 1. e3, if \ldots f5
or h5, then White will continue creating the triplet in c3--e3.
Otherwise, she is safe to create it in e3--g3. Once the Omega Weapon
is constructed, she need only move it to e2 (which is free due to the
first move, and will be adjacent to at least two pawns). This destroys
the Black king, which cannot have moved, due to the rules.
Unfortunately, this strategy can be foiled by Black with a
clever\footnote{This strategy was discovered by computer, like most
  clever things these days.} mutual suicide, as shown in
Figure~\ref{fig:chessyv2}. This strategy is not easily defended
against, nor is it straightforward to fix the rules of the
game,\!\footnote{ It seems perhaps we can eliminate Black's draw
  strategy by restoring the idea of horizontally striped Candy Rooks
  (see footnote above). I implemented this. Unfortunately, although
  the move that ends the game in a draw is a horizontal move, the move
  that creates the Candy Rook is a vertical one. Therefore, Black's
  Candy Rook is vertically striped, and can successfully destroy the
  White king (though it is then replaced with a horizontally striped
  Candy Rook to observe the empty thrones).} so it seems {\em Chessy
  Crush Saga v2} also retains hope for Black, even though the odds
initially seemed stacked against her.

\begin{figure}[ht]
\begin{center}
\chessboard[setfen=4k3/ppp1pppp/8/8/3p4/2PP4/PP2PPPP/4K3]
\end{center}\vspace{-0.1in}
\caption{ 
%
  White may not use a fixed strategy to construct her Omega Weapon.
  The position shown is after 1. c3 d5 2. d3 d4. If 3. e3 {\em (making
    candy)} then \ldots xe3, obviously. White can make candy with 3.
  b3, but now e2 is not clear to crush the Candy Rook here and destroy
  the Black king.
%
}
\label{fig:twopaths}
\end{figure}

% \chessboard[setfen=4k3/ppp1pppp/8/8/8/4p3/PP3PPP/4K3]
% \chessboard[setfen=4k3/1ppppppp/8/8/8/8/pR2PPPP/4K3]
\begin{figure}[ht]
\begin{center}
\chessboard[setfen=4k3/3ppppp/1r6/8/8/2R5/PP3PPP/4K3]
(a) \\
\chessboard[
  setfen=4k3/3ppppp/4r3/8/8/8/2R2PPP/4K3,
  pgfstyle=border,
  color=blue!50,
  markfile=e,
  color=black,
  pgfstyle=straightmove,
  markmoves={b6-e6},
  pgfstyle=cross,
  color=red,
  backfields={e8,e7,e6}]
(b)
\end{center}\vspace{-0.1in}
\caption{ 
%
  Black's surprising draw strategy in {\em Chessy Crush v2}. White
  begins her Omega Weapon: 1. e3 c6 2. d3 a6 3. c3 {\em (making
    candy)} b6 {\em (making candy)} we have the board in (a). It seems
  that Black is stuck after 4. Rc2 {\em (making candy)}, since 5. Re2
  {\em (making candy)} will crush White's candy rook and destroy the
  entire e file, including Black's king. Black defending with 4.
  \ldots Re6 initially seems pointless (White will destroy the rook as
  well so the check is irrelevant---this indeed refutes 4. \ldots Rb1)
  but Re6 actually creates a {\em vertical} three-piece crush,
  sacrificing Black's king to destroy the e file first, which also
  destroys White's king! Such mutual destruction is not possible in
  traditional chess, but it seems appropriate to consider it a draw in
  Chessy Crush Saga v2. It appears that White can head off this line
  with 4. Rc8++, a totally vanilla back-rank mate, but this is not
  actually mate with the same 4. \ldots Re6 {\em (making candy)}
  escaping to a draw.
%
}
\label{fig:chessyv2}
\end{figure}

\section{Future work and conclusions}

Many other games can be combined with Chess to ruin Chess. For example,
consider {\em Chess Tac Toe}, {\em Chesslers of Catan}, {\em Chesstris},
and {\em Hungry Hungry Hipposchess}. Combining Chess and Sokoban is
impossible, obviously. The combination of Chess and Battleship is
{\em Battlechess}, published in 1988 by Interplay.

\medskip
\medskip
In conclusion, San Dimas High School Footchess rules!

% \setchessboard{tinyboard,showmover=false}
% \chessboard[
%   setfen=rnbqkbnr/pppppppp/8/8/8/8/NNNBBNNN/QQQQKQQQ]

@@

% TODO: Use mainline to see if any more chesstego is solvable.
%
